---
title: "Dummy Variables"
output: pdf_document
---

* * * * 

A **dummy variable** or **binary variable** is a variable that takes on a value of 0 or 1 as an indicator that the observation has some kind of characteristic.  Common examples:

* Sex (female): FEMALE=1 if individual in the observation is female, equal to 0 otherwise
* Race (White): WHITE=1 if individual in the observation is white/Caucasian, equal to 0 otherwise
* Urban vs Rural: URBAN=1 if individual in the observation lives in an urban area, equal to 0 otherwise
* College graduate: COLGRAD=1 if individual in the observation has a four-year college degree, equal to 0 otherwise

It is common to use dummy variables as explanatory variables in regression models, if binary categorical variables are likely to influence the outcome variable.

## 1. Example: Factors Affecting Monthly Earnings  ##

Let us examine a data set that explores the relationship between total monthly earnings (`MonthlyEarnings`) and a number of variables on an interval scale (i.e. numeric quantities) that may influence monthly earnings including including each person's IQ (`IQ`), a measure of knowledge of their job (`Knowledge`), years of education (`YearsEdu`), and years experience (`YearsExperience`), years at current job (`Tenure`).

The data set also includes dummy variables that may explain monthly earnings, including whether or not the person is black / African American (`Black`), whether or not the person lives in a Southern U.S. state (`South`), and whether or not the person lives in an urban area (`Urban`).

The code below downloads a CSV file that includes data on the above variables from 1980 for 935 individuals and assigns it to a data set that we name `wages`.  

```{r}
wages <- read.csv("http://murraylax.org/datasets/wage2.csv");
```

The following call to `lm()` estimates a multiple regression predicting monthly earnings based on the eight explanatory variables given above, which includes three dummy variables.  The next call to `summary()` displays some summary statistics for the estimated regression.

```{r}
lmwages <- lm(MonthlyEarnings 
              ~ IQ + Knowledge + YearsEdu + YearsExperience + Tenure
              + Black + South + Urban,
              data = wages)
summary(lmwages)
```

The p-values in the right-most column reveal that all of the coefficients are statistically significantly different from zero at the 5\% significance level.  We have statistical evidence that all of these variables influence monthly earnings.

The coefficient on `Black` is equal to -110.67.  This means that even after accounting for the effects of all the other explanatory variables in the model (includes educational attainment, experience, location, knowledge, and IQ), black / African American people earn on average \$110.67 less per month than non-black people.

The coefficient on `South` is -50.82.  Accounting for the impact of all the variables in the model, people that live in Southern United States earn on average \$50.82 less per month than others.

The coefficient on `Urban` is 155.43.  Accounting for the impact of all the variables in the model, people that live in urban areas earn \$155.43 more per month, which probably reflects a higher cost of living.

We can compute confidence intervals for these effects with the following call to `confint()`
```{r}
confint(lmwages, parm=c("Black", "South", "Urban"), level = 0.95)
```

## 2. Dummy Interactions with Numeric Explanatory Variables ##

We found that black people have lower monthly earnings on average than non-black people.  In our regression equation, this implies that the *intercept* is lower for black people than non-black people.  We can also test whether a dummy variable affects the *slope* multiplying other variables.

For example, are there differences in the returns to education for black versus non-black people?  To answer this, we include an *interaction effect* between `Black` and `YearsEdu`:

```{r}
lmwages <- lm(MonthlyEarnings
              ~ IQ + Knowledge + YearsEdu + YearsExperience + Tenure
              + Black + South + Urban + Black*YearsEdu,
              data = wages)
summary(lmwages)
```

We see here that when accounting for an interaction effect between race and education, the coefficient on the `Black` dummy variable becomes insignificant, but the coefficient on the interaction term is negative and significant at the 10\% level.  The coefficient on the interaction term equal to -35.03 means the slope on education is 35.03 less when $Black=1$.

The coefficient on the interaction term is interpreted as the *additional* marginal effect of the numeric variable for the group associated with the dummy variable equal to 1.  For this example:

* The marginal effect on monthly earnings for non-black people for an additional year of education is equal to \$50.07 (i.e. when $Black=0$).

* The marginal effect on monthly earnings for black people for an additional year of education is equal to \$50.07 - \$35.03 = \$15.02 (i.e. when $Black=1$).

* Said another way, the marginal effect on monthly earnings for an additional year of education is \$35.03 less for black people than non-black people.

