---
title: "Kruskal Wallis Test for Differences in Distribution"
author: 
  - "James M. Murray, Ph.D."
  - "University of Wisconsin - La Crosse"
date: "Updated: `r format(Sys.time(), '%B %d, %Y')`"

output: 
  html_document:
    keep_md: yes
  pdf_document:
    fig_width: 7
    fig_height: 3.5
urlcolor: blue
---

[PDF file location: http://www.murraylax.org/rtutorials/kw.pdf](http://www.murraylax.org/rtutorials/kw.pdf)

[HTML file location: http://www.murraylax.org/rtutorials/kw.html](http://www.murraylax.org/rtutorials/kw.html)

* * * * 

*Note on required packages:*  The following code requires the package `psych` to perform statistics related to the median. The code also requires a feature in the `tidyverse` package.  If you have not already done so, download, install, and load the library with the following code:

`install.packages("psych")`  

`install.packages("tidyverse")`  

`library("psych")`

`library("tidyverse")`

* * *
```{r, echo=FALSE, include=FALSE}
library("psych")
library("tidyverse")
```

## 1. Introduction ##

**Kruskal Wallis Test** is a statistical method to determine if there is a difference in distributions between two or more *independent* groups, where the groups are defined by the outcomes for a single categorical variable. 

Kruskal Wallis Test can be assumed to test the hypothesis that the *center* (as measured by the median or interpolated median) of the distribution for the outcome variable is different for different categories of the explanatory variable, if it can be assumed that the *shape of the distribution* is the same across categories.  

In its purpose, it is essentially an non-parametric alternative of the one-way ANOVA for a difference in means for two or more groups.

Because the Kruskal Wallis test is a non-parametric test using ranks of the observations in its calculations, it can be performed on *ordinal* data or interval/ratio data.

## 2. Hypotheses ##

The null hypothesis says that the center of the distribution for the outcome variable all the groups are equal to each other.  The alternative hypothesis says that *at least one* of the groups has a different measure of center than the others.  

Let $K\geq 2$ denote the number of groups for the explanatory variable.  Let $M_i$ denote the population interpolated median for group $i$.  The null and alternative hypotheses for the Kruskal Wallis Test are given by,

**Null:** $M_1 = M_2 = ... = M_K$ (i.e. all the interpolated medians are equal)

**Alternative:** In at least one pair-wise comparison, $M_i \neq M_j$, for some $i\neq j$ (i.e. at least one interpolated median is different than another).

## 3. Assumptions ##
The following assumptions are necessary for the Kruskal Wallis test:

- **Random assignment:** The group that an observations belongs to is determined independently from the outcome variable.

- **Independent groups:** Observations in one group are independent of observations in all other groups.  There must therefore be different sampling units in each group.

- **Homogeneity in distribution shape:** The shape of the distribution of the outcome variable for the different groups are the same, even though the population medians or interpolated medians may be different.


## 4. Example: Current Population Survey ##

The data set below includes data from more than 52,000 individuals over the age of 25 years that participated in the 2016 Current Population Survey.  The variables include the following:

- `HOURS`: Usual weekly hours worked over all jobs (ratio)
- `SEX`: Male or female (categorical)
- `RACE`: Racial identity (categorical)
- `EDUC`: Education (categorical)
- `EDUNUM`: A numerical code for the education categories in `EDUC` where larger numbers indicate a higher level of educational attainment
- `MARRIED`: Marital status (categorical)

The following code downloads the data, opens it, and saves it in a data frame called `wdat`.

```{r}
load(url("http://murraylax.org/datasets/cpshours.RData"))
```

We can use this data to determine if educational attainment, an ordinal variable, is different for people in different categories for one or more of other the categorical variables.

## 5. Computing Medians / Interpolated Medians ##

Let us calculate the median and interpolated median level of education for the different racial groups. The code below uses the pipe operator (`%>%`) with the `group_by()` and `summarise()` functions, to compute and report the median level of education by group.

```{r}
wdat %>%
  group_by(RACE) %>%
  summarise(MedianEdu = median(EDUCNUM))
```

We can do the same thing to summarize the interpolated median by racial category.
```{r}
wdat %>%
  group_by(RACE) %>%
  summarise(MedianEdu = interp.median(EDUCNUM))
```

We can see there are differences in the sample medians and interpolated medians for level of education between different races.  In the next section, we formally test for a difference in medians.

## 6. Conducting the Kruskal Wallis Hypothesis Test ##

The code below calls the `kruskal.test()` function to test the hypothesis that average educational level is different for people in different races.

```{r}
kt <- kruskal.test(EDUCNUM ~ RACE, data=wdat)
```

The *formula*, `EDUCNUM ~ RACE` tells the function to examine the outcome variable `EDUCNUM` as it depends on `RACE`. The second parameter, `data=wdat`, specifies the data frame to find these variables.

We can review the outcome of the statistical test by viewing the return object `kt`.

```{r}
kt
```

With a p-value equal to $2.2 \times 10^{-16}$, we have sufficient statistical evidence that the average level of educational attainment is different for different racial groups.

