---
title: "Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
load(url("http://murraylax.org/datasets/facebook.RData"))
```

# Homework Assignment: Introductory Statistics #

## Introduction ##
These exercises require the `psych` packages and packages in the `tidyverse`. The `tidyverse` actually contains many packages that allow you to organize, summarize, and plot data. The package `psych` is used to perform statistics related to the median. 

The exercises below uses a data set comes from the following study on Facebook marketing and performance metrics:

Moro, S., Rita, P. and Vala, B., (2016) "Predicting Social Media Performance Metrics and Evaluation of the Impact on Brand Building: A Data Mining Approach" *Journal of Business Research*, Vol. 68, pp. 3341-3351. Available at http://www.sciencedirect.com/science/article/pii/S0148296316000813

The data frame is pre-loaded and is named `df`.

Use the `str()` function below to view the variable names and brief descriptions.
```{r varnames, exercise=TRUE}
str(df)
```

The data set includes statistics from 500 Facebook posts in 2014 related to the marketing of a globally known cosmetic brand. Facebook marketing is an important part of many businesses marketing strategy. Facebook interaction can help businesses build their brand and market new products. Marketing executives such statistics to better understand the effectiveness of their Facebook marketing.

The data set includes the following variables:

1. **Type**: Scale / Class: Nominal / Factor. Type of post. Possible outcomes are "Link", "Photo", "Status", and "Video"

2. **Month**: Scale / Class: Ordinal / Ordered factor. Month of the year for the post.

3. **Weekday**: Scale / Class: Ordinal / Ordered factor. Day of the week for the post.

4. **Hour**: Scale / Class: Ratio / Integer. Hour of the day - between 0 (12:00AM) and 23 (11:00PM)

5. **Paid**: Scale / Class: Binary / Integer. Dummy variable equal to 1 if a paid post, 0 if a free or unsolicited post.

6. **Reach**: Scale / Class: Ratio / Integer. Number of unique individuals who saw the post appear on their news feeds.

7. **Impressions**: Scale / Class: Ratio / Integer. Number of times the post appeared on people's news feeds (some individuals may have had the post appear more than once)

8. **EngagedUsers**: Scale / Class: Ratio / Integer. Number of unique individuals that clicked anywhere in the post. 

9. **Comments**: Scale / Class: Ratio / Integer. Number of comments on the post.

10. **Likes**: Scale / Class: Ratio / Integer. Number of likes for the post

11. **Shares**: Scale / Class: Ratio / Integer. Number of shares for the post

12. **Interactions**: Scale / Class: Ratio / Integer. The sum, Comments + Likes + Shares.

13. **Weekday.Int**: Scale / Class: Ordinal / Integer: Number associated with day of the week in **Weekday**

14. **Month.Int**: Scale / Class: Ordinal / Integer: Number associated with month in **Month**

## Descriptive Statistics ##

The variable, `Comments` is the number of comments for a Facebook post. Compute the mean number of comments per Facebook post.
```{r meancomments, exercise=TRUE}

```

```{r meancomments-hint}
mean(df$Comments, na.rm=TRUE)
```

```{r meancomments-check}
checker <- function(label, user_code, solution_code, check_code, envir_result, evaluate_result, ...) {
  
}
```
## Topic 1

### Exercise 

*Here's a simple exercise with an empty code chunk provided for entering the answer.*

Write the R code required to add two plus two:

```{r two-plus-two, exercise=TRUE}

```

### Exercise with Code

*Here's an exercise with some prepopulated code as well as `exercise.lines = 5` to provide a bit more initial room to work.*

Now write a function that adds any two numbers and then call it:

```{r add-function, exercise=TRUE, exercise.lines = 5}
add <- function() {
  
}
```

## Topic 2

### Exercise with Hint

*Here's an exercise where the chunk is pre-evaulated via the `exercise.eval` option (so the user can see the default output we'd like them to customize). We also add a "hint" to the correct solution via the chunk immediate below labeled `print-limit-hint`.*

Modify the following code to limit the number of rows printed to 5:

```{r print-limit, exercise=TRUE, exercise.eval=TRUE}
mtcars
```

```{r print-limit-hint}
head(mtcars)
```

### Quiz

*You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.*

Some questions to verify that you understand the purposes of various base and recommended R packages:

```{r quiz}
quiz(
  question("Which package contains functions for installing other R packages?",
    answer("base"),
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("grid", correct = TRUE)
  )
)
```

