---
title: "Dummy Variables in Regression"
subtitle: "<a href='https://murraylax.org/rtutorials' target='blank'>R Tutorials for Applied Statistics</a>"
author: 
  - name: "<a href='https://murraylax.org' style='color:white' target='blank'>James M. Murray, Ph.D.</a>"
    affiliation: "<a href='https://www.uwlax.edu' style='color:white' target='blank'>University of Wisconsin - La Crosse</a><br/><a href='https://murraylax.org/rtutorials' style='color:white; font-weight: bold;' target='blank'>R Tutorials for Applied Statistics</a>"
output:
  rmdformats::readthedown:
    highlight: haddock
    number_sections: true
---

* * * * 

A **dummy variable** or **binary variable** is a variable that takes on a value of 0 or 1 as an indicator that the observation has some kind of characteristic.  Common examples:

* Sex (female): FEMALE=1 if individual in the observation is female, equal to 0 otherwise
* Race (White): WHITE=1 if individual in the observation is white/Caucasian, equal to 0 otherwise
* Urban vs Rural: URBAN=1 if individual in the observation lives in an urban area, equal to 0 otherwise
* College graduate: COLGRAD=1 if individual in the observation has a four-year college degree, equal to 0 otherwise

It is common to use dummy variables as explanatory variables in regression models, if binary categorical variables are likely to influence the outcome variable.

# Example: Factors Affecting Monthly Earnings 

Let us examine a data set that explores the relationship between total monthly earnings (`MonthlyEarnings`) and a number of variables on an interval scale (i.e. numeric quantities) that may influence monthly earnings including including each person's IQ (`IQ`), a measure of knowledge of their job (`Knowledge`), years of education (`YearsEdu`), and years experience (`YearsExperience`), years at current job (`Tenure`).

The data set also includes dummy variables that may explain monthly earnings, including whether or not the person is black / African American (`Black`), whether or not the person lives in a Southern U.S. state (`South`), and whether or not the person lives in an urban area (`Urban`).

The code below downloads data on the above variables from 1980 for 663 individuals and assigns it to a dataframe called `df`.  

```{r}
load(url("http://murraylax.org/datasets/wage2.RData"))
```

The following call to `lm()` estimates a multiple regression predicting monthly earnings based on the eight explanatory variables given above, which includes three dummy variables.  The next call to `summary()` displays some summary statistics for the estimated regression.

```{r}
lmwages <- lm(MonthlyEarnings 
              ~ IQ + Knowledge + YearsEdu + YearsExperience + Tenure
              + Black + South + Urban,
              data = df)
summary(lmwages)
```

The p-values in the right-most column reveal that all of the coefficients are statistically significantly different from zero at the 5\% significance level.  We have statistical evidence that all of these variables influence monthly earnings.

The coefficient on `Black` is equal to -107.91. This means that even after accounting for the effects of all the other explanatory variables in the model (includes educational attainment, experience, location, knowledge, and IQ), black / African American people earn on average \$107.91 less per month than non-black people.

The coefficient on `South` is -37.84.  Accounting for the impact of all the variables in the model, people that live in Southern United States earn on average \$37.84 less per month than others.

The coefficient on `Urban` is 174.63.  Accounting for the impact of all the variables in the model, people that live in urban areas earn \$174.63 more per month, which probably reflects a higher cost of living.

We can compute confidence intervals for these effects with the following call to `confint()`
```{r}
confint(lmwages, parm=c("Black", "South", "Urban"), level = 0.95)
```

# Dummy Interactions with Numeric Explanatory Variables

We found that black people have lower monthly earnings on average than non-black people.  In our regression equation, this implies that the *intercept* is lower for black people than non-black people.  We can also test whether a dummy variable affects the *slope* multiplying other variables.

For example, are there differences in the returns to education for black versus non-black people?  To answer this, we include an *interaction effect* between `Black` and `YearsEdu`:

```{r}
lmwages <- lm(MonthlyEarnings
              ~ IQ + Knowledge + YearsEdu + YearsExperience + Tenure
              + Black + South + Urban + Black*YearsEdu,
              data = df)
summary(lmwages)
```

We see here that when accounting for an interaction effect between race and education, the coefficient on the `Black` dummy variable becomes insignificant, but the coefficient on the interaction term is negative and significant at the 10\% level.  The coefficient on the interaction term equal to -52.08 means the slope on education is 52.08 less when $Black=1$.

The coefficient on the interaction term is interpreted as the *additional* marginal effect of the numeric variable for the group associated with the dummy variable equal to 1.  For this example:

* The marginal effect on monthly earnings for non-black people for an additional year of education is equal to \$47.84 (i.e. when $Black=0$).

* The marginal effect on monthly earnings for black people for an additional year of education is equal to \$47.84 - \$52.08 = -\$4.24 (i.e. when $Black=1$), which implies a near zero and possibly negative return to education on income for the black population (you would need to test the hypothesis for the linear combination to answer this).

# Interacting Dummy Variables with Each Other

Let us interact two of the dummy variables to understand this interpretation and motivation.  In the call to `lm()` below, we use our baseline model and interact `South` and `Urban`:

```{r}
lmwages <- lm(MonthlyEarnings
              ~ IQ + Knowledge + YearsEdu + YearsExperience + Tenure
              + Black + South + Urban + South*Urban,
              data = df)
summary(lmwages)
```

To interpret the meaning of the coefficient on `South`, `Urban`, and `South*Urban`, we will ignore (hold constant) all the terms in the regression equation that do not include one of these variables.

## Difference between Urban and Rural Workers in the North/East/West

Workers in the North / East / and West U.S. have $South=0$.  Here $South=0$, $(South~x~Urban)=0$, so neither the coefficient on the interaction nor the coefficient on `South` come into play. 

The coefficient for $b_{Urban}$ implies that *in the Non-Southern U.S.*, urban workers earn on average \$213.08 more in monthly earnings than rural workers.

## Difference between Urban and Rural Workers in the South

When focusing on workers in the South, $South=1$ and the interaction term comes into play.

- Impact for urban workers in the south = $b_{South}(1) + b_{Urban}(1) + b_{Urban*South}(1)$
- Impact for rural workers in the south = $b_{South}(1) + b_{Urban}(0) + b_{Urban*South}(0)$
- Difference = $b_{Urban} + b_{Urban*South} = 213.08 - 104.97 =$ \$108.11 

*In the Southern U.S. states*, urban workers on average earn \$108.11 more in monthly earnings than rural workers.

## Difference between Southern and North/East/West Monthly Earnings for *Urban* Workers

- Impact for Southern urban workers = $b_{South}(1) + b_{Urban}(1) + b_{Urban*South}(1)$
- Impact for Non-Southern urban workers = $b_{South}(0) + b_{Urban}(1) + b_{Urban*South}(0)$
- Difference = $b_{South} + b_{Urban*South} = 34.93 - 104.97 =$ -\$70.04

*For urban workers*,  workers in the South earn \$70.04 less in monthly earnings than workers in the North/East/West.

## Difference between Southern and North/East/West Monthly Earnings for *Rural* Workers

Rural workers have $Urban=0$ and so the interaction term $Urban~x~South=0$, so we can ignore both of those coefficients.  The coefficient for $b_{South}$ implies that Southern rural workers earn on average \$34.93$ *more* per month than Non-Southern rural workers.

## Three-Way Interactions and Higher!

What?!  Things aren't complicated enough for you?!  Do at your own peril!  

I have seen people include higher order interaction effects like `South * Urban * Black * YearsEdu` in their regressions.  It has never been obvious to me that they understood what their results meant.

