---
title: "Estimating the Population Mean"
author: 
  - "James M. Murray, Ph.D."
  - "University of Wisconsin - La Crosse"
date: "Updated: `r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document
#  pdf_document
urlcolor: blue
---

[PDF file location: http://www.murraylax.org/rtutorials/one_sample.pdf](http://www.murraylax.org/rtutorials/one_sample.pdf)

[HTML file location: http://www.murraylax.org/rtutorials/one_sample.html](http://www.murraylax.org/rtutorials/one_sample.html)

* * *

The **population mean** is a measure of the center or "average" value in the whole population of a variable measured at the interval or ratio level. 

The **sample mean** is a sample estimate of the population mean.  It is a the same measure of center, obtained from a sample.  The variable in your sample must be measured at the interval or ratio level.

**Example:** Current Population Survey from 2004 that includes data on average hourly earnings, marital status, gender, and age for thousands of people.  A part of it is available for download from the textbook website for Stock and Watson's *Introduction to Econometrics*.

##1. Download the Dataset.
The code below downloads the data set and assigns the dataset to a variable we create and call `cps04`.

```{r}
cps04 <- read.csv(url("http://murraylax.org/datasets/cps04.csv"))
```

The dataset `cps04` contains a variable called `ahe`, which stands for average hourly earnings.  

##2. Compute the Sample Mean.

```{r}
mean(cps04$ahe)
```
The sample estimate for average hourly earnings for U.S. workers is 2004 is $16.77.  This is not necessarily the population mean.  Like every statistic, it includes a margin of error due to random sampling error.

##3. Compute a 95% Confidence Interval
The confidence interval is a range of values for the population mean, based on our estimate of the sample mean, and an estimate for the margin of error due to random sampling.  

The function `t.test` computes a number of statistics and statistical tests for a variable, including a confidence interval.  In the code below, we use the function to compute our  confidence interval and assign all the resulting output to a new variable we call `ahestats`.
```{r}
ahestats <- t.test(cps04$ahe, conf.level = 0.95)
```
The output of `t.test` that we assigned to variable `ahestats` is a list which includes an item called `conf.int`.  

Let's call this item to report our confidence interval:
```{r}
ahestats$conf.int
```
The confidence interval for average hourly earnings for U.S. workers is 2004 is $16.58 - $16.96.  We can say with 95% confidence that this interval estimate includes the true population mean.

##4. One Sample T-Test (One-tailed):
Suppose a politician claimed that the average earnings of American workers was more than $16.50 per hour.  We know that the sample estimate is larger from above, but let's test the hypothesis that the *population mean* is *more than* $16.50.  

The appropriate statistical procedure is the **One-sample T-test for a Mean** which tests whether a single population mean is equal to or different than a particular value.  Our null and alternative hypotheses for our one-sample t-test is given by the following:

**Null hypothesis: $\mu = 16.50$   
Alternative hypothesis: $\mu > 16.50$**

The `t.test` function can also compute the one-sample t-test using the following code:
```{r}
t.test(cps04$ahe, mu=16.50, alternative = "greater")
```
The output of the test reveals a p-value equal to 0.00028.  Since this is below 5%, we reject the null hypothesis and conclude that we do have statistical evidence that the population mean is greater than $16.50.

##5. Two-Tailed Test:
The previous example is a **one-tailed** test.  That is, it involved an alternative hypothesis that looked for statistical evidence that the population parameter was in a particular direction away from the null hypothesized value (in the case above, *greater than* the null hypothesis).

A two tailed test instead tests an alternative hypothesis that simply says the population parameter is *different than* the null hypothesized value, leaving the possibility that it may be less than or may be greater than the value.

Let's test the following two-tailed hypotheses:

**Null hypothesis: $\mu = 16.50$   
Alternative hypothesis: $\mu \neq 16.50$**

Notice the $\neq$ sign in the alternative hypothesis.  

We use the `t.test` function again to compute the one-sample t-test using the following code:
```{r}
t.test(cps04$ahe, mu=16.50, alternative="two.sided")
```
We can see from the output that the p-value is equal to 0.0057.  Since this is below 5%, we reject the null hypothesis and conclude that we do have statistical evidence that the population mean *is different than* $16.50.
